\LARGE \bf Human-Robot Interaction: Tackling the AI Challenges

### 2013-09-05 paper.tex
Rephrasing, rewrite...

This article shed light on some key decisional issues that are to be tackled
for a cognitive robot to share space and tasks with a human. We adopt a
constructive approach based on the identification and the effective
implementation of individual and collaborative skills. These cognitive
abilities cover geometric reasoning and situation assessment based essentially
on perspective-taking and affordances, management and exploitation of each
agent (human and robot) knowledge in separate cognitive models, natural
multi-modal communication, human-aware task planning, and human and robot
interleaved plan achievement.

@@

Human-robot interaction is a challenge for artificial intelligence. This field
indeed lays at the crossroad of several other domains of AI and requires to
tackle them in a holistic manner. Modeling humans and human cognition;
acquiring, representing, manipulating in a tractable way abstract knowledge;
reasoning on this knowledge to make decisions; instanciating those decisions
into physical actions in coordination with humans... Many AI techniques are
invited, from visual processing to symbolic reasoning, from task planing to
\emph{theory of mind} building, from reactive control to action recognition and
learning.

We do not tackle all these issues. This article attempts however to organize
them into a coherent challenge for artifical intelligence, and also explicit
some of the trails that have been explored on our robots and that result in a
deliberative, knowledge-focused, architecture for human-robot interaction.

Human-robot interaction requires to equip the robot with explicit reasoning on
the human and on its own capacities to achieve its tasks in a legible, if not
collaborative, way with human partners. In the following pages, we present a
robot control system which has been especially designed for a cognitive robot
which shares space and task with a human. It has been designed by adopting a
constructive approach based on effective individual and collaborative skills.
The system is comprehensive since it aims\fxnote{it *is* comprehensive, or it *aims* at being comprehensive?} at dealing with a complete set of
abilities articulated so that the robot controller is effectively able to
conduct in a flexible manner a collaborative task with a human partner.

We illustrate below how we deal with a typical human-robot interactive task
achievement and what are the abilities we claim are necessary. Section
\ref{sec:problem} proposes a typical human-robot interaction problem that can
be solved by the proposed robot controller. Section
\ref{sec:Framework} provides an overview of the robot controller and
introduces three activities which are described in \ref{sec:situ},
\ref{sec:plan} and \ref{sec:action}. Section \ref{sec:expes}
presents the structure an actual run on a real robot in face to face interaction with a
person. Finally, section~\ref{sec:soa} discusses and analyses the impact of
each of our contribution in the light of existing literature and alternative
approaches.

@@

\fxfatal{Develop/reorganize this section to organize the different problems of
HRI into one coherent challenge for AI. See how to articulate it with section
\ref{sec:psychocognitivegrounds}}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.9\columnwidth]{grounding_robot.pdf}
\caption{Robot reasoning about HRI and anticipation of human activities:
  sources of information are multi-modal dialogue, and observation of
  environment and human activity}
\label{fig:hri-dec}
\end{figure}

\paragraph{The Joint Action Challenge} Let us consider a robot which is
supposed to achieve interactive object manipulation, fetch and carry tasks and
similar tasks in a domestic environment. The problem we are dealing with here
is the following. Given:

\begin {itemize}
\item a joint goal, which has been previously established and agreed
  upon (through a process which is out of the scope of this paper),
\item the current situation, acquired through perception or
  deduction from previous perceptions, including the state of the
  environment of the robot and of the human,
\end {itemize}

@@

\subsection{Our Contributions Towards the Cognitive Robot}

@@

We have chosen to tackle the challenge from several ends: human-aware
navigation and motion planning~\cite{Mainprice2011}, situation assessment
coupled with motion planning~\cite{Mainprice2012}, projection of
``mightabilities'' that anticipates what surrounding agents may
do~\cite{Pandey2011}, and the design and deployment of a pervasive
\emph{knowledge-oriented} software architecture~\cite{Lemaignan2013}.

\section{From Human-Robot Interaction to a Deliberative Architecture}

\subsection{The Psycho-Cognitive Grounds of Human-Robot Interaction}
\label{sec:psychocognitivegrounds}

\fxfatal{write section summary of the challenges that human-robot interaction presents to robotics.}

@@

Fig.~\ref{fig|archi} gives an overview of the connections between the
deliberative components of our architecture~\cite{Alami2011}.

@@

This architecture moves away from standard layered approaches. Interactions
between components at the deliberative level are mostly bidirectional and we do
not introduce layers of abstraction amongst software components\footnote{We do
have lower-level modules to execute actions or manage sensors, but all
cognition-related modules reside at the same level.}. This is nicely
illustrated by the dialogue input processing. This component does not simply
act as an alternative perceptual input to the symbolic database; it also
actively queries previously acquired knowledge to disambiguate and validate the
newly created symbolic knowledge (see section~\ref{sect|com}).

@@

This interaction between knowledge and actions is also central to our approach
(as for any cognitive system): it is one of the activities of the
robot, actually shared between communication components (that can acquire
desires from interaction with agents, amongst other things) and an execution
controller that may decide to take an incoming desire into account to create
its own internal goals. The controller generates and manages intentions from
these goals with the help of a symbolic task planner, that also has direct
access to the knowledge base.

This activity is however not the backbone of our architecture. Other
activities are conducted in parallel, without being explicitly considered as
desires: assessment of the situation and the environment, dialogue (including
performative dialogue that can possibly change the internal state of the robot,
but does not lead to the creation of desires, like question answering or
statement assertion), various background monitoring and recognition tasks, etc.

@@

In our architecture (Fig.~\ref{fig|archi}), knowledge manipulation relies on a
\emph{semantic blackboard}: a central server (the {\sc Oro}
server~\cite{Lemaignan2010}) stores knowledge as it is produced by each of the
deliberative components. It conversely exposes a {\tt json}-based RPC API to
query the knowledge base~\cite{lemaignan2012kbapi}.

@@

At any time, the knowledge available to the robot may come from three sources:
Part of the knowledge is stored in an ontology (the {\sc OpenRobots Ontology}),
loaded when the robot initializes. This static source covers the
\emph{common-sense} knowledge of the robot, and optionally some
scenario-specific knowledge (for instance about objects that are to be
manipulated). The second part of the knowledge is acquired at runtime from
perception, interaction, decision making.  We go over the processes it involves
in the next section. Lastly, the third source of knowledge is the output of the
reasoner.

@@

This lead to a bidirectional design process: from \emph{bottom-up} regarding
the choices of concepts to model, \emph{top-down} regarding the upper part of the
taxonomy. This upper part of the ontology is pictured on
figure~\ref{fig|upper_tbox}. All the classes visible on this figure belong to the
{\sc OpenCyc} namespace (the {\tt cyc:} prefix is omitted for clarity).

By \emph{aligning} the upper part of the ontology on {\sc OpenCyc} (as other
KRS, like {\sc KnowRob}~\cite{Tenorth2009a} or PEIS K\&R~\cite{Daoutis2009}, did)
has multiple advantages. First the design of this part of the ontology is
generally difficult: it pertains to abstract concepts whose mutual relations
comes to philosophical debates. The upper taxonomy of {\sc OpenCyc} represents
a relative consensus, at least within the semantic Web community. Then, because
it is a well established project with numerous links to other on-line databases
(like Wikipedia or WordNet), the reuse of important {\sc OpenCyc} concepts
ensures to a certain extend that the knowledge stored by the robot can be
shared or extended with well-defined semantics. A good example is the concept
of \emph{Object}: In everyday conversation, an object is a relatively small
physical thing, that can be typically manipulated. Normally, a human is not
considered as an object. In {\sc Cyc}, an object has a more precise definition:
it is something \emph{partially tangible}. That includes obviously the humans,
and actually many other entities that would not be commonly said to be objects
(the Earth for instance). Thus the importance of relying on well-defined and
standard semantics to exchange informations between artificial systems.

This figure~\ref{fig|upper_tbox} also illustrates the fundamental disjunction
in the ORO model between \emph{temporal} and \emph{spatial} entities (formally,
$(TemporalThing \sqcap SpatialThing)^{\mathcal{I}} = \emptyset$, with
$\mathcal{I}$ the \emph{interpretation} of our model).

@@

The tree of figure~\ref{fig|upper_tbox} (this subset of the ontology is indeed
a tree: this has however not to be the case in general, and, as a matter of
fact, the TBox of the whole ORO common-sense ontology does not form a tree) is
not equally developed at lower levels. We have already briefly mentioned the
developments of the actions. The other important part is the descendants of the
\concept{partially tangible thing} (what is commonly called an \emph{object}).
Figure~\ref{fig|tangible_things_tbox} gives more details on this part of the
ontology.

@@

This excerpt from the ontology makes the bottom-up design process visible: only
few types of \emph{partially tangible} things appear, and only subclasses
relevant to the context of service robotics in an human-like environment are
present.

@@

The {\sc Oro} server also features a mechanism to mimic minimalistic forms of
biological memory.  When new statements are inserted in the knowledge base, a
\emph{memory profile} is optionally attached to them.

Three such profiles are predefined: {\tt short term}, {\tt episodic} and {\tt
long term}. They each correspond to a different lifetime for the statements
(respectively 10 seconds, 5 minutes and no time limit). After this duration,
the statements are automatically removed from the knowledge base.

This approach is limited. In particular, \emph{episodic} memory primarily
refers to the semantics of the statements (that is expected to be related to an
event) and not to a specific life duration.

We rely however on this short term memory for a particular use-case:
\emph{active concepts}. Some modules, like the natural language processor, use
the {\tt short term} memory profile to mark for a few seconds important
concepts that are currently manipulated by the robot. For example, if a human
asks the robot: ``Give me all red objects'', the human, the \concept{Give}
action, and each red objects that are found are successively marked as
\emph{active concepts} by inserting statements such as \stmt{human type
ActiveConcept} in the short-term memory (which can be considered, in this case,
to be a working memory). We use this feature to trace certain knowledge-related
processes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Implementing the Cognitive Skills}

@@

\begin{figure}
        \centering
        \includegraphics[width=\columnwidth]{spark-oro}
    \caption{Functional overview of knowledge base ({\sc Oro} server, top part) and the geometric situation assessment module (\emph{{\sc Spark}}, bottom part)}
        \label{fig|spark-oro}
\end{figure}

Anchoring perceptions in a symbolic model requires perception abilities and
their symbolic interpretation. We rely on a dedicated geometric and temporal
reasoning module called {\sc Spark} (\emph{SPAtial Reasoning \&
Knowledge}~\cite{Sisbot2011}). It is a situation assessment reasoner that
generates symbolic knowledge from the geometry of the environment with respect
to relations between objects, robots and humans (Fig.~\ref{fig|spark-oro}),
also taking into account the different perspective that each agent has on the
environment.

@@

\section{Experimental outcomes}

@@

>>>>

This article sheds light on some key decisional issues that are to be tackled
for a cognitive robot to share space and tasks with a human, and present our
take on these challenges. We adopt a constructive approach based on the
identification and the effective implementation of individual and collaborative
skills. These cognitive abilities cover geometric reasoning and situation
assessment mainly based on perspective-taking and affordances, management and
exploitation of each agent (human and robot) knowledge in separate cognitive
models, natural multi-modal communication, human-aware task planning, and human
and robot interleaved plan achievement.

@@

Human-robot interaction is a challenge for artificial intelligence. This field
indeed lays at the crossroad of several other domains of AI and requires to
tackle them in a holistic manner: Modeling humans and human cognition;
acquiring, representing, manipulating in a tractable way abstract knowledge;
reasoning on this knowledge to make decisions; and eventually instanciating
those decisions into physical actions in coordination with humans. Many AI
techniques are invited, from visual processing to symbolic reasoning, from task
planing to \emph{theory of mind} building, from reactive control to action
recognition and learning.

We do not tackle all these issues. This article attempts however to organize
them into a coherent challenge for artifical intelligence, and also explicit
some of the paths that have been investigated on our robots and that result in a
deliberative, knowledge-oriented, architecture for human-robot interaction.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.9\columnwidth]{grounding_robot.pdf}
\caption{Robot reasoning about HRI and anticipation of human activities.
  Sources of information are multi-modal dialogue, and monitoring of the
  environment and human activity. The robot must adapt on-line its behaviours
  by merging computed plans with reactive control.}
\label{fig:hri-dec}
\end{figure}

\fxfatal{Merge the following 3 paragraphs}
Our frame is the robot which is
expected to achieve interactive object manipulation, fetch and carry tasks and
similar tasks in a domestic environment (Figure~\ref{fig:hri-dec}).

@@

The problem we are dealing with here
is the following. Given:

\begin {itemize}
\item a joint goal, which has been previously established and agreed
  upon (through a process which is out of the scope of this paper),
\item the current situation, acquired through perception or
  deduction from previous perceptions, including the state of the
  environment of the robot and of the human,
\item [\it a priori} common-sense knowledge,
\end {itemize}

@@

\subsection{Our contributions towards the cognitive robot}

@@

\subsection{Article organization}

The remaining of the article discuss the robotic architecture we have built to
tackle the AI challenges of human-robot interaction, and how its relates to
other approaches. We propose to organize this discussion in four sections.

The next section presents the overall architecture and the knowledge model we
have developed for our robots. Section~\ref{sec:impl} gives some insights on
the each of the main components of the architecture, and try to highlight their
significance for artificial intelligence. Section~\ref{sec:expe} presents two
experiments that illustrate in a concrete way what can be currently achieved
with our robots.  Section~\ref{sect|conclusion} finally summarizes our main
contribution and relate them to other recent researches in the field.

\section{From human-robot interaction to a deliberative architecture}

@@

\fxfatal{Describe archi}

Figure~\ref{fig|archi} gives an overview of the connections between the
deliberative components of our architecture~\cite{Alami2011}.

@@

This architecture moves away from standard layered approaches found in
robotics~\cite{Gat1998three, Volpe2001CLARAty, Goldberg2002}. Interactions
between components at the deliberative level are mostly bidirectional and we do
not introduce layers of abstraction amongst software components\footnote{We do
have lower-level modules to execute actions or manage sensors, but all
cognition-related modules reside at the same level.}. Dialogue processing, for
instance, illustrates this structure: This component does not simply act as an
alternative perceptual input to the symbolic database; it also
actively queries previously acquired knowledge to disambiguate and validate the
newly created symbolic knowledge (see section~\ref{sect|com}).

@@

This interaction between knowledge and actions is also central to our approach
(as for any cognitive system): it is one amongst all the activities carried by
the robot, actually performed by the communication components (that acquire
desires from interaction with agents) and the execution controller that may
decide to take an incoming desire into account to create its own internal
goals. The controller generates and manages intentions from these goals with
the help of a symbolic task planner, that also has direct access to the
knowledge base.

This activity is however not the backbone of our architecture. Other
activities are conducted in parallel, without being explicitly considered as
desires: assessment of the situation and the environment, dialogue (including
performative dialogue that can possibly change the internal state of the robot,
but does not lead to the creation of desires, like statement assertion or
question answering), various background monitoring and recognition tasks, etc.

@@

In this architecture, knowledge manipulation relies on a
semantic \emph{blackboard}: a central server (the {\sc Oro}
server~\cite{Lemaignan2010}) stores knowledge as it is produced by each of the
deliberative components. It conversely exposes a {\tt json}-based RPC
API~\cite{lemaignan2012kbapi} to query the knowledge base.

@@

At any time, the knowledge available to the robot may come from three sources:
{\it a priori} knowledge is stored in an ontology (the {\sc OpenRobots Ontology}),
loaded when the robot initializes. This static source covers the
\emph{common-sense} knowledge of the robot, and optionally some
scenario-specific knowledge (for instance about objects that are to be
manipulated). The second part of the knowledge is acquired at runtime from
perception, interaction, decision making.  We go over the processes it involves
in the next sections. Lastly, the third source of knowledge is the output of the
reasoner.

@@

This lead to a bidirectional design process: from \emph{bottom-up} regarding
the choices of concepts to model, \emph{top-down} regarding the upper part of the
taxonomy. This upper part of the ontology is pictured on
Figure~\ref{fig|upper_tbox}. All the classes visible on this figure belong to the
{\sc OpenCyc} namespace (the {\tt cyc:} prefix is omitted for clarity).

Aligning the upper part of the ontology on {\sc OpenCyc} (as done by other
KRS, like {\sc KnowRob}~\cite{Tenorth2009a} or PEIS K\&R~\cite{Daoutis2009})
has multiple advantages. First the design of this part of the ontology is
generally difficult: it pertains to abstract concepts whose mutual relations
comes to philosophical debates. The upper taxonomy of {\sc OpenCyc} represents
a relative consensus, at least within the semantic Web community. Then, because
it is a well established project with numerous links to other on-line databases
(like Wikipedia or WordNet), the reuse of important {\sc OpenCyc} concepts
ensures to a certain extend that the knowledge stored by the robot can be
shared or extended with well-defined semantics. A good example is the concept
of \emph{Object}: In everyday conversation, an object is a relatively small
physical thing, that can be typically manipulated. Normally, a human is not
considered as an object. In {\sc Cyc}, an object has a more precise definition:
it is something \emph{partially tangible}. That includes obviously the humans,
and actually many other entities that would not be commonly said to be objects
(the Earth for instance). Thus the importance of relying on well-defined and
standard semantics to exchange informations between artificial systems.

Figure~\ref{fig|upper_tbox} also illustrates the fundamental disjunction
in the ORO model between \emph{temporal} and \emph{spatial} entities (formally,
$(TemporalThing \sqcap SpatialThing)^{\mathcal{I}} = \emptyset$, with
$\mathcal{I}$ the \emph{interpretation} of our model).

@@

The tree in Figure~\ref{fig|upper_tbox} (this subset of the ontology is indeed
a tree: this has however not to be the case in general, and, as a matter of
fact, the TBox of the whole ORO common-sense ontology does not form a tree) is
not equally developed at lower levels. We have already briefly mentioned the
developments of the actions. The other important part is the descendants of the
\concept{partially tangible thing} (what is commonly called an \emph{object}).
Figure~\ref{fig|tangible_things_tbox} gives more details on this part of the
ontology.

@@

This excerpt from the ontology makes the bottom-up design process visible: only
few types of \emph{partially tangible} things appear, and only subclasses
relevant to the context of service robotics in an human-like environment are
present. For performances and ease of use, we do not attempt to comprehensively
cover the field. We opportunistically extend the common-sense knowledge when
required by experiments.

@@

The {\sc Oro} server also features a mechanism to mimic minimalistic forms of
memory families.  When new statements are inserted in the knowledge base, a
\emph{memory profile} is optionally attached to them.

Three such profiles are predefined: {\tt short term}, {\tt episodic} and {\tt
long term}. They are currently attached to different lifetime for the statements
(respectively 10 seconds, 5 minutes and no time limit). After this duration,
the statements are automatically removed from the knowledge base.

This approach is limited. In particular, \emph{episodic} memory should primarily
refer to the semantics of the statements (that is expected to be related to an
event) and not to a specific life duration.

We rely however on this mechanism in certain cases: for instance, some modules
like the natural language processor use the {\tt short term} memory profile to
mark for a few seconds important concepts that are currently manipulated by the
robot as \emph{active concepts}. For example, if a human asks the robot: ``Give
me all red objects'', the human, the \concept{Give} action, and each red
objects that are found are successively marked as \emph{active concepts} by
inserting statements such as \stmt{human type ActiveConcept} in the short-term
memory (which can be considered, in this case, to be a working memory). We use
this feature to trace certain knowledge-related processes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Implementing the Cognitive Skills}
\label{sec:impl}

@@

\begin{figure}
        \centering
        \includegraphics[width=\columnwidth]{spark-oro}
    \caption{Functional overview of knowledge base ({\sc Oro} server, top part)
    and the geometric situation assessment module (\emph{{\sc Spark}}, bottom
    part). {\sc Spark} computes symbolic relationships between objects and
    agents, and fed them to the knowledge base.}

        \label{fig|spark-oro}
\end{figure}

Anchoring perceptions in a symbolic model requires perception abilities and
their symbolic interpretation. We rely on a dedicated geometric and temporal
reasoning module called {\sc Spark} (\emph{SPAtial Reasoning \&
Knowledge}~\cite{Sisbot2011}). It is a situation assessment reasoner that
generates symbolic knowledge from the geometry of the environment with respect
to relations between objects, robots and humans (Figure~\ref{fig|spark-oro}),
also taking into account the different perspective that each agent has on the
environment.

@@

\section{Experimental outcomes}
\label{sec:expe}

@@

\fxfatal{Rewrite conclusion}
